### 2017-08-18 Blog/09-ClassicalUncertaintyRelationship.tex
Reviewed blog posts

Suppose we have a phase space distribution $\rho(x,p)$. As the state evolves into its final state, the amount of material remains the same: $\int \rho \, dx \wedge dp = 1$. The final entropy will also remain the same: $I(\rho) = \int \rho \ln \rho \, dx \wedge dp = I_0$. We now ask the following question: once we fix the entropy, what is the distribution that minimizes the product $\sigma_x\sigma_p$ of the standard deviations? That would correspond to the minimum uncertainty for the final distribution given by Hamiltonian evolution on the initial distribution.

We are going to use Lagrange multipliers, which is a standard technique in calculus of variations. We write a function that is the product of the variances $\sigma_x^2 \sigma_p^2 \equiv \int (x-\mu_x)^2 \rho \, dxdp \int (p-\mu_p)^2 \rho \, dxdp$ summed with the two constraints (the total amount of material is $1$ and the total entropy $I_0$) each multiplied by a constant. Then take the derivative and set it to $0$.
\begin{align*}
L = &\sigma_x^2 \sigma_p^2 + \lambda_1\left(\int \rho dxdp - 1\right) + \lambda_2\left(- \int \rho \ln \rho \, dxdp - I_0\right)\\
\delta L = & \delta \sigma_x^2 \sigma_p^2 + \sigma_x^2 \delta \sigma_p^2 +  \lambda_1 \, \delta \int \rho dxdp - \lambda_2 \, \delta \int \rho \ln \rho \, dxdp \\
= &\int  (x-\mu_x)^2 \delta \rho \, dxdp \, \sigma_p^2 +  \sigma_x^2 \int (p-\mu_p)^2 \delta \rho \, dxdp + \\ &\lambda_1 \int \delta\rho \, dxdp - \lambda_2 \int (\delta \rho \ln \rho + \rho \delta \ln \rho)dxdp  \\
= &\int [(x-\mu_x)^2 \sigma_p^2 + \sigma_x^2 (p-\mu_p)^2 + \lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] \delta \rho  \, dxdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (x-\mu_x)^2 \sigma_p^2 + \sigma_x^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(x-\mu_x)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_x^2 (p-\mu_p)^2}{\lambda_2}}\\
L = &\int (x-\mu_x)^2 \rho \, dxdp \int (p-\mu_p)^2 \rho \, dxdp \\
&+ \lambda_1\left(\int \rho dxdp - 1\right) + \lambda_2\left(- \int \rho \ln \rho \, dxdp - I_0\right)\\
\end{align*}
Now that we found the form of the function, we solve the multipliers. That is: we find $\lambda_1$ and $\lambda_2$ such that the distribution is unitary and the information entropy is $I_0$. We have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_x \sigma_p} e^{-\frac{(x-\mu_x)^2}{2\sigma_x^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I_0 = &\ln (2\pi\sigma_x\sigma_p) + 1
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

>>>>

Suppose we have a phase space distribution $\rho(x,p)$. As the state evolves into its final state, the amount of material remains the same: $\int \rho \, dx \wedge dp = 1$. The final entropy will also remain the same: $I(\rho) = - \int \rho \ln \rho \, dx \wedge dp = I_0$. We now ask the following question: once we fix the entropy, what is the distribution that minimizes the product $\sigma_x\sigma_p$ of the standard deviations? That would correspond to the minimum uncertainty for the final distribution given by Hamiltonian evolution on the initial distribution.

We are going to use Lagrange multipliers, which is a standard technique in calculus of variations. We write a function $M$ that is the product of the variances $\sigma_x^2 \sigma_p^2 \equiv \int (x-\mu_x)^2 \rho \, dxdp \int (p-\mu_p)^2 \rho \, dxdp$ summed with the two constraints (the total amount of material is $1$ and the total entropy $I_0$) each multiplied by a constant. Then take the derivative and set it to $0$.
\begin{align*}
M = &\sigma_x^2 \sigma_p^2 + \lambda_1\left(\int \rho dxdp - 1\right) + \lambda_2\left(- \int \rho \ln \rho \, dxdp - I_0\right)\\
\delta M = & \delta \sigma_x^2 \sigma_p^2 + \sigma_x^2 \delta \sigma_p^2 +  \lambda_1 \, \delta \int \rho dxdp - \lambda_2 \, \delta \int \rho \ln \rho \, dxdp \\
= &\int  (x-\mu_x)^2 \delta \rho \, dxdp \, \sigma_p^2 +  \sigma_x^2 \int (p-\mu_p)^2 \delta \rho \, dxdp + \\ &\lambda_1 \int \delta\rho \, dxdp - \lambda_2 \int (\delta \rho \ln \rho + \rho \delta \ln \rho)dxdp  \\
= &\int [(x-\mu_x)^2 \sigma_p^2 + \sigma_x^2 (p-\mu_p)^2 + \lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] \delta \rho  \, dxdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (x-\mu_x)^2 \sigma_p^2 + \sigma_x^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(x-\mu_x)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_x^2 (p-\mu_p)^2}{\lambda_2}}
\end{align*}
Now that we found the form of the function, we solve the multipliers. That is: we find $\lambda_1$ and $\lambda_2$ such that the distribution is unitary and the information entropy is $I_0$. We have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_x \sigma_p} e^{-\frac{(x-\mu_x)^2}{2\sigma_x^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I_0 = &\ln (2\pi\sigma_x\sigma_p) + 1
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

### 2017-08-18 Blog/10-GalileanHamiltonianMechanicsDoesntReallyExist.tex
Reviewed blog posts

\section{Galilean Hamiltonian mechanics doesn't really exist}

or "Non-relativistic Hamiltonian mechanics doesn't really exist"
or "All Hamiltonian mechanics is relativistic"

TL;DR Galilean transformations are not compatible with standard Hamiltonian mechanics. They give good approximations only for small changes in velocity.

As part of my work in foundations of physics, I am forced to constantly search for different ways to look at things. Sometimes this leads me to have a different perspective which brings a more consistent overall picture but forces one to reshuffle some of the basic concepts. The relationship between relativity and Hamiltonian mechanics is one of those cases.

When studying physics, you learn that relativistic mechanics is more ``correct" and non-relativistic mechanics is just an approximation. But you may also get a sense that non-relativistic mechanics is self-consistent: it just so happens that the world is different. If you look at things closely, though, I don't think that's the case.

@@

Suppose we have a Hamiltonian system. Its state is given by position $x$ and momentum $p$. Its evolution is given by:
\begin{equation}
\begin{aligned}
\frac{dx^i}{dt} = \frac{\partial H}{\partial p_i}  \\
\frac{dp_i}{dt} = - \frac{\partial H}{\partial x^i}  \\
\end{aligned}
\label{Hamilton}
\end{equation}
where $H(x^i,p_j)$ is the Hamiltonian.

Suppose we change coordinates $\hat{x}^i = \hat{x}^i(x^j)$: how does the momentum change? We can recover the transformation rule of momentum by requiring that, in the new coordinates, the system is still Hamiltonian. One way of doing that is requiring that area elements of phase space are conserved. We find that the product of two matrices $\frac{\partial \hat{x}^i}{\partial x^j}$ and $\frac{\partial \hat{p}_i}{\partial p_k}$ must be the identity matrix, which means one is the inverse of the other.
\begin{equation*}
\begin{aligned}
dx^i \wedge dp_i = d\hat{x}^i\wedge d\hat{p}_i &=  \frac{\partial \hat{x}^i}{\partial x^j} dx^j \wedge \frac{\partial \hat{p}_i}{\partial p_k} dp_k \\
\frac{\partial \hat{x}^i}{\partial x^j} \frac{\partial \hat{p}_i}{\partial p_k} &= \delta^k_j \\
\frac{\partial \hat{p}_i}{\partial p_k} = \left[\frac{\partial \hat{x}^i}{\partial x^k}\right]^{-1} &=  \frac{\partial x^k}{\partial \hat{x}^i}
\end{aligned}
\end{equation*}

This means that $\frac{\partial \hat{p}_i}{\partial p_k}$ is only a function of position, therefore the new momentum must be a linear transformation. Integrating we have:
\begin{equation*}
\begin{aligned}
\hat{x}^i &= \hat{x}^i(x^j) \\
\hat{p}^i &= \frac{\partial x^j}{\partial \hat{x}^i} p_j + \frac{\partial f(x^j)}{\partial \hat{x}^i}
\end{aligned}
\end{equation*}
where $f$ is an arbitrary function. Since it is arbitrary, we set it to zero and have:
\begin{equation}
\begin{aligned}
\hat{x}^i &= \hat{x}^i(x^j) \\
\hat{p}^i &= \frac{\partial x^j}{\partial \hat{x}^i} p_j
\end{aligned}
\label{spaceTransf}
\end{equation}
Technically, we haven't shown why the arbitrary function added to the momentum is actually the gradient of a scalar function. Since we are setting it to zero, I didn't want to put details that are not that relevant.

>>>>

\section{Galilean transformations are not compatible with Hamiltonian mechanics}

TL;DR Galilean boosts do not leave Hamilton's equations and phase space volumes unchanged. They give good approximations only for small changes in velocity.

When studying physics, you learn that relativistic mechanics is more ``correct" and non-relativistic mechanics is just an approximation. But you may also get a sense that non-relativistic mechanics is self-consistent: it just so happens that the world is different. If you look at things closely, though, I don't think that's the case. The problem is that typically in textbooks a small sleight of hand is performed, so we have to look from a slightly different angle to see the trick.

@@

Suppose we have a Hamiltonian system. Its state is given by position $x^i$ and momentum $p_j$. Its evolution is given by:
\begin{equation}
\begin{aligned}
\frac{dx^i}{dt} = \frac{\partial H}{\partial p_i}  \\
\frac{dp_j}{dt} = - \frac{\partial H}{\partial x^j}  \\
\end{aligned}
\label{Hamilton}
\end{equation}
where $H(x^i,p_j)$ is the Hamiltonian.

Suppose we change coordinates $\hat{x}^i = \hat{x}^i(x^j)$: how does the momentum change? We can recover the transformation rule of momentum by requiring that, in the new coordinates, the system is still Hamiltonian. One way of doing that is requiring that area elements of phase space are conserved. We find that the product of the two matrices $\frac{\partial \hat{x}^i}{\partial x^j}$ and $\frac{\partial \hat{p}_i}{\partial p_k}$ must be the identity matrix, which means one is the inverse of the other.
\begin{equation*}
\begin{aligned}
dx^i \wedge dp_i = d\hat{x}^i\wedge d\hat{p}_i &=  \frac{\partial \hat{x}^i}{\partial x^j} dx^j \wedge \frac{\partial \hat{p}_i}{\partial p_k} dp_k \\
\frac{\partial \hat{x}^i}{\partial x^j} \frac{\partial \hat{p}_i}{\partial p_k} &= \delta^k_j \\
\frac{\partial \hat{p}_i}{\partial p_k} = \left[\frac{\partial \hat{x}^i}{\partial x^k}\right]^{-1} &=  \frac{\partial x^k}{\partial \hat{x}^i}
\end{aligned}
\end{equation*}

This means that $\frac{\partial \hat{p}_i}{\partial p_k}$ is only a function of position, therefore the new momentum must be a linear transformation with respect to the old momentum. Integrating we have:
\begin{equation*}
\begin{aligned}
\hat{x}^i &= \hat{x}^i(x^j) \\
\hat{p}_i &= \frac{\partial x^j}{\partial \hat{x}^i} p_j + \frac{\partial f(x^j)}{\partial \hat{x}^i}
\end{aligned}
\end{equation*}
where $f$ is an arbitrary function. Since it is arbitrary, we set it to zero and have:
\begin{equation}
\begin{aligned}
\hat{x}^i &= \hat{x}^i(x^j) \\
\hat{p}_i &= \frac{\partial x^j}{\partial \hat{x}^i} p_j
\end{aligned}
\label{spaceTransf}
\end{equation}
Technically, we haven't shown why the arbitrary function added to the momentum is actually the gradient of a scalar function. Since we are setting it to zero, I didn't want to put details that are not that relevant.
